Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-29 paper/paper.tex
Updated Results and Future work

\section{Experimental Evaluation}
\label{sec:experiments}
  Our benchmark mostly consists  of programs from nofib~\cite{nofib}
suite.  All our
benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
Ubuntu 14.04.

@@

The statistics related to compile  time liveness analysis are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this compile-time
overhead, we have added a caching mechanism in our simulator that can
reuse the DFA from earlier analysis if the program has not changed..

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of garbage collection  events,
average  number of  cells reclaimed  per garbage collection,  average number  of cells
touched  per garbage collection,  total  number of  dragged cells  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
lower)  for LGC  based collection.   LGC  also reduces  the number  of
dragged cells significantly. However, the cost  for LGC is that per garbage collection
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within 5X of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthetic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.

\newcommand{\hgt}{2.9cm}
\begin{figure*}[t]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
%\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  
%% Column (a) shows complete  usage, column 
%%    (b) enlarges  a part  of the  usage.  
The blue and the red curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The black curve represents the number of
reachable cells and the  lightblue curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}

Memory  usage graphs  for select  benchmarks are  shown
Figure~\ref{fig:memory-usage},  Column   (a).   As  the
number of  garbage collections  tend to be  very large,
for  each benchmark  we  also show  a  window which  is
representative  of  the  behavior for  that  particular
benchmark (Column (b)).  In all the programs we can see
that  the   curve  corresponding  to  LGC   (red  line)
regularly dips below the RGC curve (blue line), the dip
is  specially  visible  for the  benchmarks  where  LGC
outperforms RGC substantially.  The graphs also include
curve    for   reachable    cells   (black,    obtained
approximately  by forcing  RGC to  run at  a very  high
frequency)  and  the  live  cells  for  particular  run
(lightblue, obtained by post processing the heap access
at the end of a program).

One area of concern is  the huge gap between the actual
liveness and  the liveness perceived by  our collector.
In  case of  LGC for  eager languages~\cite{asati14lgc}
the gap was  very narrow and almost  touched the actual
liveness curve.  In case of lazy languages, this gap is
largely   due   to   approximate   reachability   based
collection  of closures.  To  implement liveness  based
collection of closures, we need to record liveness data
for  each  closure and  update  it  at each  evaluation
point.   These space  and  run time  overheads make  it
infeasible  to   incorporate  liveness  based   garbage collection  for
closures in a practical garbage collectors.

@@

%% One issue we highlighted  was that while fewer nodes
%% were marked (and  more garbage collected), sometimes
%% cons cells  could be visited and  traversed multiple
%% times  with  different  sets of  liveness  paths  to
%% explore.  Further work  includes improvements to the
%% classical copying  collector to  reduce the  cost of
%% this.
We have  extended liveness-based garbage collection  to lazy languages
and shown its benefit for  standard benchmark programs.  This was done
by defining a liveness analysis of heap objects that we proved correct
with  respect   to  a   non-standard  semantics   that  served   as  a
specification of  liveness.  The result  of the  analysis is a  set of
grammars, whose membership question was  proved to be undecidable. The
grammars are  therefore approximated by  DFAs and used by  the garbage
collector to improve collection.

@@

\warning{Although we  provide an  implementation for  a liveness
based garbage  collection scheme for lazy  language, we
do not  provide a formal  proof of its  correctness.  A
formal proof of correctness would describe how closures
are handled  during garbage collection and  liveness is
propagated inside closures.

  Another  interesting exercise  is to  narrow the  gap
  between  the   actual  liveness  and   the  perceived
  liveness  of our  garbage collector.  Our experiments
  (not reported here) show that a significant number of
  dead  cells  get   trapped  inside  closures.   Using
  strictness  analysis  to  eagerly  evaluate  closures
  might release more of these  dead cells to be garbage
  collected.   Since our  garbage collector  can handle
  closures   (suspended  evaluations),   another  major
  challenge  is to  extend it  to support  higher order
  programs.

>>>>

\section{Experimental Evaluation}
\label{sec:experiments}
  Our benchmark consists  of programs taken from nofib~\cite{nofib}
and other similar fucntional programming test suites.  All our
benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
Ubuntu 14.04.

@@

The statistics related to liveness analysis and DFA generation are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this 
overhead, we use a caching mechanism in our simulator that can
reuse the DFA from earlier analysis if the program has not changed.

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of garbage collection  events,
average  number of  cells reclaimed  per garbage collection,  average number  of cells
touched  per garbage collection,  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
lower)  for LGC  based collection.   However, the cost  for LGC is that per garbage collection
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within \warning{5X} of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthetic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.

\newcommand{\hgt}{2.9cm}
\newcommand{\LGCLine}{red}
\newcommand{\RGCLine}{blue}
\newcommand{\ReachLine}{black}
\newcommand{\UseLine}{lightblue}

\begin{figure*}[t]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
%\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  
%% Column (a) shows complete  usage, column 
%%    (b) enlarges  a part  of the  usage.  
The \RGCLine\ and the \LGCLine\ curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The \ReachLine\ curve represents the number of
reachable cells and the  \UseLine\ curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}

Memory  usage graphs  for select  benchmarks are  shown
Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
number of  garbage collections  tend to be  very large,
hence  we  show only  a  slice of the original graph  which  is
representative  of  the  behavior for  that  particular
benchmark (Column (b)).  In all the programs we can see
that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
regularly dips below the RGC curve (\RGCLine\ line), the dip
is  specially  visible  for the  benchmarks  where  LGC
outperforms RGC substantially.  The graphs also include the
curve    for   reachable    cells   (\ReachLine,    obtained
approximately  by forcing  RGC to  run at  a very  high
frequency)  and  the  live  cells  for a particular run
(\UseLine, obtained by post processing the heap access
at the end of the program).

As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
but still falls short of the precision achieved by LGC in case of eager languages
~\cite{asati14lgc}. 
A case for concern is the huge difference in the number of cells touched by LGC vs RGC for some programs.
This is due to heap locations being  traversed multiple times during LGC. \warning{We have implemented heuristics to 
overcome common cases and they show promising results}.

@@

%% One issue we highlighted  was that while fewer nodes
%% were marked (and  more garbage collected), sometimes
%% cons cells  could be visited and  traversed multiple
%% times  with  different  sets of  liveness  paths  to
%% explore.  Further work  includes improvements to the
%% classical copying  collector to  reduce the  cost of
%% this.
We have  extended liveness-based garbage collection  to lazy languages
and shown its benefit for a set of benchmark programs.  This was done
by defining a liveness analysis of heap objects that we proved correct
with  respect   to  a   non-standard  semantics   that  served   as  a
specification of  liveness.  The result  of the  analysis is a  set of
grammars, whose membership question was  proved to be undecidable. The
grammars are  therefore approximated by  DFAs and used by  the garbage
collector to improve collection.

@@

\warning{Although we  provide an  implementation for  a liveness
based garbage  collection scheme for a lazy  language, we
do not  provide a formal  proof of its  correctness.  A
formal proof of correctness would describe how closures
are handled  during garbage collection and  liveness is
propagated inside closures.

  Another  interesting exercise  is to  narrow the  gap
  between  the   actual  liveness  and   the  perceived
  liveness  of our  garbage collector. Using
  strictness  analysis  to  eagerly  evaluate  closures
  might release more dead cells to be garbage
  collected.  Since our  garbage collector  can handle
  closures   (suspended  evaluations),   another  major
  challenge  is to  extend it  to support  higher order
  programs.
